Substation Analysis February 22, 2016
===============================================================================

## Goal
The goal of this report is to quickly find Substations which are performing 
outside of the required limits and provide the ability for analysis for 
February 22, 2016.

```{r Load Packages, include=FALSE}
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(knitr)
library(RColorBrewer)

#install.packages("lubridate")
library(lubridate)
```

### Variables for the limits for the charts.  

> voltageHigh = 128  
  voltageLow  = 114  
  pfHigh      = 0.99  
  pfLow       = 0.98  

### Upper limit of voltage.  

> voltageHighLimit = 140  
  Any voltage above 140v will be considerd invalid and removed from the analysis.  
  
```{r ConstantVariables, include=FALSE}

# Setup variables for the limits for the charts
voltageHigh = 128
voltageLow  = 114

pfHigh = 0.99
pfLow  = 0.98

# Upper limit of voltage.  Any voltage above 140v
# will be considerd invalid and removed from the analysis
voltageHighLimit = 140

```

```{r Functions, include=FALSE}
# Trim leading and trailing whitespaces
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
```

```{r Load_the_Data, include=FALSE}
# Set Working Directory
setwd("/Users/cn/rcode/substation/")

# Read date is in local time for both datasets
v <- read.csv("voltage.csv")
pf <- read.csv("pf.csv")

# data dictionary for u
# voltage.csv  (3 vars)
# 1. ReadDate - Local Time as factor, 96 intervals for the day
# which in utlitlity terms for electric meter is 15 minute readings
# a reading every 15 minutes.  
# 2. Voltage the voltage measured on the meter at the specific time
# period.
# 3. substation name

# pf.csv (powerfactor)
# 1. Read Date 
# 2. ReadValue - hourly data - needs to be cleaned, since the data
# is offset by 10minutes for each hour and we need match this
# value to the hour for the 
# 3. station -> needs to be mapped to substationName
# SCADA.Point (mega watts, mega vars delivered, mega vars received)
# 
# 
#
#
```

```{r RemoveDuplicatesFromData, include=FALSE}

# remove duplicates for voltage and power factor datasets
v_dedupe <- v  %>% 
  group_by(readdate,Voltage, substationName) %>% 
  filter(row_number() == 1)

pf_dedupe <- pf  %>% 
  group_by(ReadValue, station, SCADA.Point, ReadDate, substationName) %>% 
  filter(row_number() == 1)

```


```{r Format_the_DataPowerFactor, include=FALSE}
# Ppivot the data for the pf calculation
# SCADA.Point contains, megawatts, megavars delievered, and megaVars received
pf_wide <- spread(pf_dedupe,SCADA.Point, ReadValue)

# rename the columns
pf_wide <- rename(pf_wide, mw=PMWD3D, mvar.delivered=PMQD3D, mvar.received=PMQD3R)

# Date Formating
# v$dt <- format(as.POSIXct(v$readdate, format="%m/%d/%y %H:%M"), format="%Y-%m-%d")
# v$hm <- format(as.POSIXct(v$readdate, format="%m/%d/%y %H:%M"), format="%H:%M")
# pf$dt <- format(as.POSIXct(pf$ReadDate, format="%m/%d/%y %H:%M"), format="%Y-%m-%d")
# pf$m <- format(as.POSIXct(pf$ReadDate, format="%m/%d/%y %H:%M"), format="%M")
# v$m <- format(as.POSIXct(v$readdate, format="%m/%d/%y %H:%M"), format="%M")
# Add and hour field to the voltage data.  The data is interval ending. This means
# the dates need to be shifted 15 minutes back if we use the transformation below
# to avoid associating the first 3 intervals for the day with the last interval for
# the day (which is the next day at zero hundred hours (00:00))
# One way to solve this is convert readdate to as.Date( ) and then subtract 15mins.
# Subtracting 15mins from the date will only work for intervals less than one hour
#pf_wide$dtReadDate <- parse_date_time(x, "%m/%d/%y %H:%M",tz = "America/Chicago")
pf_wide$dtReadDate <- parse_date_time(pf_wide$ReadDate, orders="mdy hm")
pf_wide$dtReadDate  <- pf_wide$dtReadDate  - minutes(10)
pf_wide$h <- hour(pf_wide$dtReadDate)

#pf_wide$h <- format(as.POSIXct(pf_wide$ReadDate, format="%m/%d/%y %H:%M"), format="%H")

# force hour numeric, we will do the same for v_dedupe
# so we can join these columns on each other to join the 2 datasets
pf_wide$h <- as.numeric(pf_wide$h)

# calculate the total MVARs
# PMQD3R (mvar.received) can be .na, when this occurs just use the mvar.delievered
# See references, regarding Chapel Hill and for calculating power factor.
# Set mvar to the difference (deliver-received, if no received then delievered) 
pf_wide$mvar <- ifelse(!is.na(pf_wide$mvar.received), 
                       (pf_wide$mvar.delivered-pf_wide$mvar.received) 
                       ,pf_wide$mvar.delivered)

# calculate the power factor using diff or Delieverd - Received
# Power Factor = kw / (kw^2 + (Delieverd-Received)^2)  
# Power Factor = mw / (mw^2 + (mvar)^2)  
pf_wide$pf <- pf_wide$mw / sqrt((pf_wide$mw * pf_wide$mw + pf_wide$mvar * pf_wide$mvar))

# calculate the power factor as normal, for review and comparison
# Power Factor = kw / (kw^2 + kvar^2) 
#pf_wide$pfB <- pf_wide$PMWD3D / sqrt((pf_wide$PMWD3D*pf_wide$PMWD3D + pf_wide$PMQD3D*pf_wide$PMQD3D))

# add the leading and lagging field to pf_wide and maybe... totalsummary and outliers...
## If mvar (+) Lagging, If mvar (-) Leading
## I THINK I SHOULD MAKE ALL IF ELESE THE SAME SO SWAP TO THIS TO MATCH +1/-1 correction
# pf_wide$desc <- ifelse(pf_wide$mvar < 0, 'Leading' , 'Lagging')
pf_wide$desc <- ifelse(pf_wide$mvar < 0, 'Leading' , 'Lagging')

# right shift the leading power factors for the chart
# chart center is 1, and to show the directrion of the 
# phase angle we shift values to the right for the leading power factor
# TODO Figure out if leading is to the right on chart...!!!!!!!!!  If not swap direction of sign (>)
pf_wide$pfChart <- ifelse( pf_wide$mvar < 0, (1-pf_wide$pf)+1 ,pf_wide$pf)

```

```{r Format_the_Data_Voltage, include=FALSE}

# clean and roll back 15min to accomodate interval ending and hours
v_dedupe$dtReadDate <- parse_date_time(v_dedupe$readdate, orders="mdy hm")
v_dedupe$dtReadDate  <- v_dedupe$dtReadDate  - minutes(15)
v_dedupe$h <- hour(v_dedupe$dtReadDate)
# old school posixct for x axis lables
v_dedupe$hm <- format(as.POSIXct(v_dedupe$readdate, format="%m/%d/%y %H:%M"), format="%H:%M")

# force the value to numeric like we have for pf_wide$h
v_dedupe$h <- as.numeric(v_dedupe$h)

# divide all the votlages by 2 since we have 240v want to see 120v instead
# requirement: from customer
v_dedupe$Voltage <- v_dedupe$Voltage / 2

# Remove bad values - anything over 180v, right, ... 
# TODO: Add reference 
#v_dedupe = v_dedupe[v_dedupe$Voltage < 180,]
v_dedupe <- subset(v_dedupe, v_dedupe$Voltage < voltageHighLimit)

# Summarize the data for this dataset by Substation
# to be used in a table for review.
voltageBySubstation <- v_dedupe %>%
  group_by(substationName)  %>%
  summarise(mean = mean(Voltage),
            median = median(as.numeric(Voltage)),
            min = min(Voltage),
            max = max(Voltage),
            n = n())  %>% 
  arrange(substationName)

```

```{r Merge_the_Data, include=FALSE}

# create new dataframes with just the data we need to review
# to prepare to merge the two datasets so we can plot the bivarate charts
pf_col <- select(pf_wide,substationName,h,mvar ,pf, pfChart)
v_col <- select(v_dedupe, readdate,substationName,h, hm, Voltage)

# clean up the white spaces and set all substationNames to Upper Case

pf_col$substationName <- trim(pf_col$substationName)
v_col$substationName <- trim(v_col$substationName)

# set all names to UCase
pf_col$substationName <- toupper(pf_col$substationName)
v_col$substationName <- toupper(v_col$substationName)

# merge the data
# we have the same pf for each 4 intervals for a 15min reading
# since pf is hourly and voltage is by 15min
# we can review this later... we just need to see these values
# all grouped up and displayed 
total <- left_join(v_col, pf_col, by= c("substationName", "h"))

```

```{r Total_the_Data, include=FALSE}

# create a summary of the merged data for voltage and power factor
# totalSummary is for each substation, including n, the count of records
totalSummary <- total %>%
  group_by(substationName)  %>%
  summarise(mean = mean(Voltage),
            vmedian = median(as.numeric(Voltage)),
            vmin = min(Voltage),
            vmax = max(Voltage),
            pfmean = mean(pf),
            pfmedian = median(as.numeric(pf)),
            pfmin = min(pf),
            pfmax = max(pf), 
            mvars = sum(mvar),
            n = n())  %>% 
  arrange(substationName)

# Setup a flag to monitor the direction of the mvars (lagging & leading)
# For records with shift > 0 we need to create a new pf value so we can
# chart the values as right shifted from the pf = 1.  pfChart is used
# to hold the new value.  The goal here is to have values which can be charted
# to follow this x-axis points (.94,.96,.98,1,.98,.96,.94) with the values to
# the left of 1 being lagging and values to the right of 1 being leading
# If mvar (+) Lagging, If mvar (-) Leading
# FIGURE OUT THE SHIFT ON THE CHART
totalSummary$pfChart <- ifelse( totalSummary$mvar < 0, (1-totalSummary$pfmedian)+1 ,totalSummary$pfmedian)

# THIS IS DEAD
outliers <- subset(totalSummary,
                   totalSummary$pfmedian <0.98 |
                   totalSummary$pfmedian>0.99 |
                   totalSummary$vmedian>126 |
                    totalSummary$vmedian<116.5 )


# Leading and Lagging Analysis
# Divide the lagging and leading dataframes into individual 
# data frames to help the user review the data in tables
# RENAME... as neg as RIGHT SHIFTED, and _pos LEFT SHIFTED
# OUR LEADING AND LAGGING
leading <- subset(totalSummary, totalSummary$mvars <  0)
lagging <- subset(totalSummary, totalSummary$mvars >  0)

# TODO, add back the negative$pfmedian <0.96 and negative$pfmedian > 0.98
# or use the pfChart value b/c values for outlier_pos have been right
# shifted from the 1 (1 is center for this chart, see references #Lagging and Leading)
# RENAME... as neg as RIGHT SHIFTED, and _pos LEFT SHIFTED
leading_outliers <- subset(leading, 
                    leading$pfmedian < pfLow |
                    leading$pfmean > 1.02 |
                    leading$vmedian < voltageLow |
                    leading$vmedian > voltageHigh) 
                   
# since our true value for pf is in a range(s)
#0.98 to 1.02 max, the lagging will never cross over 1.0 b/c of the power factor calculation.
#So I can remove this from the filter... 
lagging_outliers <- subset(lagging, 
                   lagging$pfmedian < pfLow |
                   #lagging$pfmedian > pfHigh  |
                   lagging$vmedian  < voltageLow  |
                   lagging$vmedian  > voltageHigh )

# Tag the values for easy review
totalSummary$desc <- ifelse(totalSummary$mvar < 0, 'Lead' , 'Lag')

```

```{r SecondAttemptForMinMaxPowerFactor,include=FALSE}
# Second Attempt calcualte the min max from shifted values
totalSummary2 <- total %>%
  group_by(substationName)  %>%
  summarise(mean = mean(Voltage),
            vmedian = median(as.numeric(Voltage)),
            vmin = min(Voltage),
            vmax = max(Voltage),
            pfmean = mean(pfChart),
            pfmedian = median(as.numeric(pfChart)),
            pfmin = min(pfChart),
            pfmax = max(pfChart), 
            mvars = sum(mvar),
            n = n())  %>% 
  arrange(substationName)

# Fix all the leading power factors where pf* > 1
# we will shift this values back to their original value
totalSummary_pretty <- totalSummary2
totalSummary_pretty$pfmeanC <- ifelse(totalSummary_pretty$pfmean > 1, 
                                      (1-totalSummary_pretty$pfmean)+1, totalSummary_pretty$pfmean)

totalSummary_pretty$pfmedianC <- ifelse(totalSummary_pretty$pfmedian > 1, 
                                      (1-totalSummary_pretty$pfmedian)+1, totalSummary_pretty$pfmedian)

totalSummary_pretty$pfminC <- ifelse(totalSummary_pretty$pfmin > 1, 
                                      (1-totalSummary_pretty$pfmin)+1, totalSummary_pretty$pfmin)

totalSummary_pretty$pfmaxC <- ifelse(totalSummary_pretty$pfmax > 1, 
                                      (1-totalSummary_pretty$pfmax)+1, totalSummary_pretty$pfmax)


# Leading and Lagging Analysis
# Divide the lagging and leading dataframes into individual 
# data frames to help the user review the data in tables
# RENAME... as neg as RIGHT SHIFTED, and _pos LEFT SHIFTED
# OUR LEADING AND LAGGING
leading_p <- subset(totalSummary_pretty, totalSummary_pretty$mvars <  0)
lagging_p <- subset(totalSummary_pretty, totalSummary_pretty$mvars >  0)

# TODO, add back the negative$pfmedian <0.96 and negative$pfmedian > 0.98
# or use the pfChart value b/c values for outlier_pos have been right
# shifted from the 1 (1 is center for this chart, see references #Lagging and Leading)
# RENAME... as neg as RIGHT SHIFTED, and _pos LEFT SHIFTED
leading_outliers_p <- subset(leading_p, 
                    leading_p$pfmedian < pfLow |
                    leading_p$pfmedian > 1.02 |
                    leading_p$vmedian < voltageLow |
                    leading_p$vmedian > voltageHigh) 
                   
# since our true value for pf is in a range(s)
#0.98 to 1.02 max, the lagging will never cross over 1.0 b/c of the power factor calculation.
#So I can remove this from the filter... 
lagging_outliers_p <- subset(lagging_p, 
                   lagging_p$pfmedian < pfLow |
                   #lagging$pfmedian > pfHigh  |
                   lagging_p$vmedian  < voltageLow  |
                   lagging_p$vmedian  > voltageHigh )


```

## Voltage Lagging and Leading Daily

```{r Daily Substation Voltage & Power Factor,echo=FALSE,fig.width=12, fig.height=12, echo=FALSE, warning=FALSE }

tsh <-  subset(totalSummary_pretty, totalSummary_pretty$substationName=='HILLSBORO')
tsh <-  subset(totalSummary_pretty, !is.na(totalSummary_pretty$pfmedian))

chart1 <- ggplot(data=tsh[!is.na(tsh$pfmedian), ], 
       aes(x=pfmedian, y=vmedian)) +
    geom_point(size=3, color='green') + 
    geom_point(data=leading_outliers_p, aes(x=pfmedian, y=vmedian), color='red', size=4) +
    geom_text(data=leading_outliers_p, aes(label=substationName),hjust=0, vjust=2,size=3) +
    geom_errorbar(data=leading_outliers_p, aes(ymin = vmin, ymax = vmax), width = 0.0025, alpha=1/2) + 
    geom_point(data=lagging_outliers_p, aes(x=pfmedian, y=vmedian), color='red', size=4) +
    geom_text(data=lagging_outliers_p, aes(label=substationName),hjust=0, vjust=2,size=3) +
    geom_errorbar(data=lagging_outliers_p, aes(ymin = vmin, ymax = vmax), width = 0.0025, alpha=1/2) + 
    #geom_text(aes(label=substationName),hjust=-0.2, vjust=0.5,size=3) +
    #geom_errorbar(aes(ymin = vmin, ymax = vmax), width = 0.0025,alpha=1/2) + 
    #pf is RIGHT shifted for Leading values so I would need to shift the max and min as well &*&^!
    #geom_errorbarh(aes(xmin = pfmin, xmax = pfmax,  height = .4),alpha=1/2) +
    geom_errorbarh(data=leading_outliers_p, aes(xmax = pfmax, xmin = pfmin, height = .4), alpha=1/2) +
    geom_errorbarh(data=lagging_outliers_p, aes(xmax = pfmax, xmin = pfmin, height = .4), alpha=1/2) +
    scale_y_continuous(limits = c(110, 130)) +
    scale_x_continuous(limits = c(0.92, 1.10),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, yend = voltageHigh), size=0.15,colour='black') + # left
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, yend = voltageHigh), size=0.15,colour='black') + # right
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, yend = voltageHigh), size=0.05,colour='black') + # top
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, yend = voltageLow), size=0.15,colour='black') + # bottom
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica')  +
    ggtitle("Substation Voltage & Power Factor" )

# remove all the outliers from the main dataset
tsh <-  subset(totalSummary_pretty, !is.na(totalSummary_pretty$pfmedian))
tsh <-  tsh[!(tsh$substationName %in% leading_outliers_p$substationName),]
tsh <-  tsh[!(tsh$substationName %in% lagging_outliers_p$substationName),]

chart2 <- ggplot(data=tsh[!is.na(tsh$pfmedian), ], 
       aes(x=pfmedian, y=vmedian,color=substationName)) +
    geom_point(size=3) + 
    geom_text(aes(label=substationName),hjust=-0.2, vjust=0.5,size=3) +
    geom_errorbar(aes(ymin = vmin, ymax = vmax), width = 0.0025,alpha=1/2) + 
    #pf is RIGHT shifted for Leading values so I would need to shift the max and min as well &*&^!
    geom_errorbarh(aes(xmin = pfmin, xmax = pfmax,  height = .4),alpha=1/2) +
    scale_y_continuous(limits = c(110, 130)) +
    scale_x_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, yend = voltageHigh), size=0.15,colour='black') + # left
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, yend = voltageHigh), size=0.15,colour='black') + # right
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, yend = voltageHigh), size=0.05,colour='black') + # top
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, yend = voltageLow), size=0.15,colour='black') + # bottom
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica') +
    #labs(linetype='custom title') +
    #scale_fill_continuous(guide = guide_legend()) +
    theme(legend.position="bottom")  +
    ggtitle("Substation Voltage & Power Factor" )

grid.arrange(chart1, arrangeGrob(chart2), ncol = 1)

```

Outliers Description:  
* Voltage Outlier (median) which will be above or below the high/low voltage  
* Power factor (median) which will left or right shift the substation on the chart.  

We do have to match the power factor to the hour and need to review if repeating the power factor for each interval in the hour avoids introduces anomolies. 

## Boxplot of Power Factor

```{r BoxPlotOfPowerFactor, echo=FALSE,fig.width=12, fig.height=10, echo=FALSE, warning=FALSE }
# reshape the y axis to fit the power factor use case / user expectations
ggplot(data=pf_wide[!is.na(pf_wide$pfChart), ], 
       aes(factor(substationName), pfChart)) +
    geom_boxplot(outlier.colour = "red") + 
    scale_y_continuous(limits = c(0.90, 1.10),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    ylab("power factor") + 
    geom_hline(yintercept = 1, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Power Factor by Substation" )

# calculate the boxplot's data for output
# http://www.statmethods.net/management/aggregate.html
pfboxplotdata <- pf_wide %>% 
  group_by(substationName) %>% 
  summarise(min=min(pf,na.rm = TRUE), 
            Q1=quantile(pf, probs=0.25,na.rm = TRUE), 
            median=median(pf,na.rm = TRUE),
            mean=mean(pf,na.rm = TRUE),
            Q3=quantile(pf, probs=0.75,na.rm = TRUE),
            max=max(pf,na.rm = TRUE))  %>% 
        arrange(substationName)
```

### Boxplot Power Factor By Substation Data

```{r BoxPlotPowerFactorData, echo=FALSE}

kable(pfboxplotdata, digits=3)

```


This chart is used to show the shape of the distribution, its central value, and its variability. In a box and whisker plot: the ends of the box are the upper and lower quartiles, so the box spans the interquartile range. the median is marked by a vertical line inside the box

The upper and lower "hinges" correspond to the first and third quartiles (the 25th and 75th percentiles). 

The upper whisker extends from the hinge to the highest value that is within 1.5 * IQR of the hinge, where IQR is the inter-quartile range, or distance between the first and third quartiles. The lower whisker extends from the hinge to the lowest value within 1.5 * IQR of the hinge. Data beyond the end of the whiskers are outliers and plotted as points (as specified by Tukey).

* Outliers are in red are values outside of the 3rd quartile.

## Histograms (Voltage & Power Factor)

```{r HistogramVoltagePowerFactor, echo=FALSE,fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}
# histogram of voltage
p2<- ggplot(data = total,aes(x = Voltage)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram (bin=0.5)")

# Histogram of power factor by lagging & leading
# need to avoid overlaying the bars on each other
p3<- ggplot(data=pf_wide[!is.na(pf_wide$pfChart), ]
            ,aes(x = pfChart, fill=desc)) + 
  geom_histogram(binwidth=0.002, color = 'black') + 
  theme_bw(base_family= 'Helvetica') + 
  theme(legend.position="bottom") + 
  theme(legend.position="none") +
  scale_x_continuous(limits = c(0.92, 1.08),
                       breaks=c(.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c("0.92",".94", ".96", ".98","1", ".98", ".96",".94","0.92")) +
  xlab("power factor") +
  ggtitle("Power Factor Histogram (bin=0.002)")

grid.arrange(p2, arrangeGrob(p3), ncol = 2)

# TODO:  Greg mentioned the original intent for this chart was to see the voltage and pf in read time
#        and i need my notes to figure out what green, yellow, and red mean again... 
```

A histogram is a bar graph of raw data that creates a picture of the data distribution. The bars represent the frequency of occurrence by classes of data. A histogram shows basic information about the data set, such as central location , width of spread , and shape.

### Lagging Substations 

```{r OutlierKTableNeg, echo=FALSE}

kable(lagging_outliers, digits=3)

```

### Leading Substations 

```{r OutlierKTablePos, echo=FALSE}

kable(leading_outliers, digits=3)

```

### All Substations

Need to pick a preference on what to display for the min and max and how to distinguish when the min is lagging and max is leading, or when the substation min and max spans the distance between the lagging and leading power factor.

```{r KTabletotalSummary, echo=FALSE}

kable(totalSummary_pretty, digits=2)

```

This table has multiple columns used to help right shift the charts and also the fields ending in 'C' which is the corrected value which is the 'original' value of the power factor without the leading shift.

Substations missing from the chart are missing data, either data for voltage or data needed to calculate power factor.  

TODO: An additional historical report of the daily min-max review of power factor might be usefull when reporting leading and lagging at the daily level.  Or the number of times in the day when the substation switches from leading to lagging or lagging to leading.

## Voltage by 15min Intervals

```{r Voltage15MinuteIntervalsByHour, fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}

#TODO NEED x axis labels
ggplot(data=total, aes(x=hm, y=Voltage)) +
  geom_point(aes(color=substationName),position=position_jitter(h=1),alpha=1/2, size=2) + 
  scale_y_continuous(limits = c(100, 130)) +
  scale_fill_brewer(palette="Set2") +
  geom_hline(yintercept=voltageLow) +
  geom_hline(yintercept=voltageHigh) +
  ggtitle("Voltage by 15min Intervals Scatter Plot (Raw Data - no dupes)" ) +
  theme_bw(base_family= 'Helvetica') +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
  #theme(axis.text.x = element_text(angle = -90, hjust = 0,size=6)) 

```

### Summary of Voltage for all data points.

```{r SummaryOfVoltageData, echo=FALSE, warning=FALSE}
#remove_max <- subset(total, total$Voltage < max(total$Voltage))
summary(total$Voltage)
```

```{r VoltageBuckets, fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}
total$voltage.bucket <- cut(total$Voltage, 
                             breaks = c(0,110,115,120,125,130))

ggplot(data = total, aes(x=hm, y=Voltage))+
  geom_point(aes(color=voltage.bucket), size=2, position='jitter', alpha=.75)+
  scale_y_continuous()+
  scale_color_brewer(type="seq") +
  ggtitle("Voltage Intervals Grouped") +
  xlab("hm")+
  ylab("voltage") +
  #theme(axis.ticks = element_blank(), axis.text.x = element_blank())
  theme(axis.text.x = element_blank())

# ggplot(aes(x = hm, y = Voltage),
#        data = subset(total, !is.na(voltage.bucket))) +
#        geom_line(aes(color=voltage.bucket), stat='summary', fun.y=median)
```

### Summary of Voltage Buckets
```{r SummaryOfVoltageBucketData, echo=FALSE, warning=FALSE}
#remove_max <- subset(total, total$Voltage < max(total$Voltage))
summary(total$voltage.bucket)
```

Summary of Voltage for all data points excluding the max value for Lynchburg.

TODO: Another suggestion is that we can build groups of ranges and the number of incidents in those ranges is the factor
to sort and search.

## Boxplot of Voltage

```{r BoxPlotOfVoltage, echo=FALSE,fig.width=12, fig.height=10, echo=FALSE, warning=FALSE }
# reshape the y axis to fit the power factor use case / user expectations
ggplot(data=total[!is.na(total$Voltage), ], 
       aes(factor(substationName), Voltage)) +
    geom_boxplot(outlier.colour = "red") + 
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    geom_hline(yintercept = voltageHigh, color='blue') +
    geom_hline(yintercept = voltageLow, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Voltage by Substation" ) 

# http://www.statmethods.net/management/aggregate.html
vboxplotdata <- total %>% 
  group_by(substationName) %>% 
  summarise(min=min(Voltage,na.rm = TRUE), 
            Q1=quantile(Voltage, probs=0.25,na.rm = TRUE), 
            median=median(Voltage,na.rm = TRUE),
            mean=mean(Voltage,na.rm = TRUE),
            Q3=quantile(Voltage, probs=0.75,na.rm = TRUE),
            max=max(Voltage,na.rm = TRUE))  %>% 
        arrange(substationName)
```

### Boxplot Power Factor By Substation Data

```{r BoxPlotVoltageData, echo=FALSE}

kable(vboxplotdata, digits=3)

```


This chart is used to show the shape of the distribution, its central value, and its variability. In a box and whisker plot: the ends of the box are the upper and lower quartiles, so the box spans the interquartile range. the median is marked by a vertical line inside the box

The upper and lower "hinges" correspond to the first and third quartiles (the 25th and 75th percentiles). 

The upper whisker extends from the hinge to the highest value that is within 1.5 * IQR of the hinge, where IQR is the inter-quartile range, or distance between the first and third quartiles. The lower whisker extends from the hinge to the lowest value within 1.5 * IQR of the hinge. Data beyond the end of the whiskers are outliers and plotted as points (as specified by Tukey).

* Outliers are in red.

## Voltage & Power Factor Substation Outliers
Goal here is to see the movement of the substation voltage and power factor on the corrected chart displayed first in this analysis to see how the data moves in the regions for each individual interval of data.

## Substation Voltage & Power Factor Intervals

```{r SubstationVoltagePowerFactorIntervals, echo=FALSE,fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}

ggplot(data=total[!is.na(total$pfChart), ], 
       aes(x=pfChart, y=Voltage)) +
    geom_point(aes(color=substationName),position=position_jitter(h=1),alpha=1/2, size=2) + 
    scale_fill_brewer(palette="Set2") +  
    scale_y_continuous(limits = c(100, 130)) +
    scale_x_continuous(limits = c(0.94, 1.06),
                       breaks=c(.94,.96,.98,1,1.02,1.04,1.06),labels=c(".94", ".96", ".98","1", ".98", ".96",".94")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, yend = voltageHigh), size=0.15,colour='black') + # left
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, yend = voltageHigh), size=0.15,colour='black') + # right
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, yend = voltageHigh), size=0.05,colour='black') + # top
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, yend = voltageLow), size=0.15,colour='black') + # bottom
    ggtitle("Substation Voltage & Power Factor" ) +
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica') 

```

This chart represents the voltage and power factor for all data points collected for February 22, 2016. The chart allows one to see how much of the data falls out of the voltage and power factor defined region (0.98 (left) to 0.98 (right), and 114v to 128v).

We can see we have more points with Lagging power factors outside the required range for efficient power factor values.

## Histogram of Voltage Distribution (All Substations)

```{r HistogramOfVoltage, fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}

h1 <- ggplot(data = total,aes(x = Voltage)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram by Substation\n bin=0.5")

h2 <- ggplot(data = subset(total,total$Voltage<117) ,aes(x = Voltage)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram by Substation (v < 117)\n bin=0.5")

h3 <- ggplot(data = subset(total,total$Voltage>126) ,aes(x = Voltage),  binwidth=0.05) + 
  geom_histogram(binwidth=0.25, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram by Substation (v > 126)\n bin=0.25")

grid.arrange(h1, arrangeGrob(h2, h3), ncol = 2)

```

A histogram is a bar graph of raw data that creates a picture of the data distribution. The bars represent the frequency of occurrence by classes of data. A histogram shows basic information about the data set, such as central location , width of spread , and shape.

## Substation Voltage Histograms
```{r VoltageHistchart, fig.width=10, fig.height=12, echo=FALSE, warning=FALSE}

ggplot(data = v_dedupe,aes(x = Voltage)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  facet_wrap(~ substationName, scales='free', ncol = 3) +
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram by Substation (bin=0.5)")

```

A histogram is a bar graph of raw data that creates a picture of the data distribution. The bars represent the frequency of occurrence by classes of data. A histogram shows basic information about the data set, such as central location , width of spread , and shape.

Would it be benefical to have hitogram by day for a single substation over a time period?  Maybe histogram per month...


## Substation Voltage by Time (intervals)

```{r VoltageScatterchart, fig.width=10, fig.height=14, echo=FALSE, warning=FALSE}

ggplot(data=v_dedupe, aes(x=hm, y=Voltage)) +
  geom_point(alpha=1/5) + 
  facet_wrap(~ substationName, ncol = 3) +
  theme_bw(base_family= 'Helvetica')  +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())

```

The scatter plot shows the data along the time axis for the intervals for the day. The interesting point in this chart, which is similar to the histogram is how the shading changes from dark to grey, which is the points stacking on top of each other.  

It takes 5 points on top of each other to make a solid point on this chart.  This demonstrates how the data is spread out over the ranges through the day by hour.

### Voltage Summary Data

```{r SummaryVoltageBySub, echo=FALSE}
kable(voltageBySubstation, digits=2)
```

## Additional Charts for exploratory review.

### Power Factor by Substation and hour 

```{r PowerFactorChart voltage, fig.width=10, fig.height=8,echo=FALSE, warning=FALSE}

ggplot(aes(x = h, y = pf, color=desc), data = pf_wide) + 
  geom_point(size = 2, position = 'jitter') +
  facet_wrap(~substationName) + 
  ggtitle('Power Factor (Leading and Lagging)') +
  theme_bw(base_family= 'Helvetica') 

```

Leading is colored instead of right shifted in these charts.

### Power Factor Summary By Substation

```{r PowerFactorSummarySort voltage, echo=FALSE}

#sort the data for display since we are reviewing data by hour
sorted_pf <- pf_wide[order(pf_wide$substationName, pf_wide$h),] 
kable(sorted_pf, digits=2)

```

### MegaWatt By Substation

```{r MeagWattsScatterPlot, fig.width=10, fig.height=8, echo=FALSE, warning=FALSE}

ggplot(data=pf_wide[!is.na(pf_wide$mw), ], aes(x=h, y=mw)) +
  geom_point() + 
  facet_wrap(~substationName, scales='free') +
  ggtitle("Mega Watt Scatter Plot by Substation" ) +
  theme_bw(base_family= 'Helvetica') 

```

### MegaVARS By Substation

```{r MegaVarsScatterPlot, fig.width=10, fig.height=8, echo=FALSE, warning=FALSE}

mvars <- subset(pf_dedupe, pf_dedupe$SCADA.Point == 'PMQD3D' | pf_dedupe$SCADA.Point == 'PMQD3R')

ggplot(data=mvars,aes(x=ReadDate, y=ReadValue, color=SCADA.Point)) +
  geom_point() +
  facet_wrap(~substationName, scales='free', ncol=3) +
  ggtitle("Mega VARs Scatter Plot by Substation" ) +
  theme_bw(base_family= 'Helvetica')  +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())

```

## References

### Bad Values
Any voltage over 180v is considered bad data, or a spike and the data will be removed.

### Interval Ending
Votlage Data

### Chapel Hill Power Factor 
Chapel Hill has a different power meter.  It does not register MVARS received in a separate register.  It is (+) if delivered and (–) if received.  Therefore, pf will have to be calculated with just the one number for MVARS. 

### Votlage Calculation
Voltage is being divided by 2 to adjust the view to 120 volts since the data being collected is 240 volts.

### Power Factor  
Instructions regarding PF calculation:  
https://en.wikipedia.org/wiki/Power_factor  

http://www.slideshare.net/RossWelshVA/power-factor-correction-the-easiest-biggest-green-initiative

Q, Reactive Power - kVAR -> PMQD3D Delivered, PMQR3D Received  
  (PMQD3D-PMQR3D)  total VARs  
S, Apparent Power    
P, Real Power - kW -> PMWD3D (MW)

S^2 = P^2 + Q^2  
S   = SQRT(p^2 + Q^2)  

Power Factor = (P, Real Power) / (S, Apparent Power)  
Power Factor = kw / (kw^2 + kvar^2)  
Power Factor = kw / (kw^2 + (Delieverd-Received)^2)  

### Would it be nice to see the trend of kw or kvar increasing or decreasing to predict when power factor is on a downward trend.... 

### Leading and Lagging Power Factor
* Delivered vars - Received vars   > 0 Lagging  
* Delivered vars - Received vars   < 0 Leading

Summary of notes to be reviewed:

The original definition of MAX PF:  

Least positive Lagging PF value for the given set of data.    

A Lagging PF is defined as any value where PMQD3D – PMQD3R is greater than zero.  Because PMQD3D is all zeroes for BOYNTON and CULLEOKA for the samples below, there is never a Lagging PF value for the date.  Therefore, the MAX PF would be zero for the given day.  

From the discussion:   
· Min PF - Least negative (Leading) PF value for the given subset  
· Max PF - Least positive (Lagging) PF value for the given subset  
· Avg PF - Most common value for the given subset OR Mean Value  

Notes:
Phase 1 - is just charting and quick review of the data.  
Phase 2 - will apply statistical models and analysis if required.

TODO: This is a daily analysis so the data is available after the day is over.  To improve this
type of analysis we need to see patterns and trends which occur which can help predict an event which is about to happen. 