Substation Analysis February 22, 2016
===============================================================================

## Goal
The goal of this report is to quickly find Substations which are performing 
outside of the required limits and provide the ability for analysis for 
February 22, 2016.

```{r Load Packages, include=FALSE}
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(knitr)
library(RColorBrewer)

#install.packages("lubridate")
library(lubridate)
```

### Variables for the limits for the charts.  

> voltageHigh = 128  
  voltageLow  = 114  
  pfHigh      = 0.99  
  pfLow       = 0.98  

### Upper limit of voltage.  

> voltageHighLimit = 140  
  Any voltage above 140v will be considerd invalid and removed from the analysis.  
  
```{r ConstantVariables, include=FALSE}

# Setup variables for the limits for the charts
voltageHigh = 128
voltageLow  = 114

pfHigh = 0.99
pfLow  = 0.98

# Upper limit of voltage.  Any voltage above 140v
# will be considerd invalid and removed from the analysis
voltageHighLimit = 140

```

```{r Functions, include=FALSE}
# Trim leading and trailing whitespaces
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
```

```{r Load_the_Data, include=FALSE}
# Set Working Directory
setwd("/Users/cn/rcode/substation/")

# Read date is in local time for both datasets
v <- read.csv("voltage.csv")
pf <- read.csv("pf.csv")

# data dictionary for u
# voltage.csv  (3 vars)
# 1. ReadDate - Local Time as factor, 96 intervals for the day
# which in utlitlity terms for electric meter is 15 minute readings
# a reading every 15 minutes.  
# 2. Voltage the voltage measured on the meter at the specific time
# period.
# 3. substation name

# pf.csv (powerfactor)
# 1. Read Date 
# 2. ReadValue - hourly data - needs to be cleaned, since the data
# is offset by 10minutes for each hour and we need match this
# value to the hour for the 
# 3. station -> needs to be mapped to substationName
# SCADA.Point (mega watts, mega vars delivered, mega vars received)
# 4. Leadings vs Lagging - Direction - is this categorical????
# Along with the Ranges too!
# 
#
#
```

```{r RemoveDuplicatesFromData, include=FALSE}
# First two tidy datasets for the raw data
# remove duplicates for voltage and power factor datasets
v_dedupe <- v  %>% 
  group_by(readdate,Voltage, substationName) %>% 
  filter(row_number() == 1)

pf_dedupe <- pf  %>% 
  group_by(ReadValue, station, SCADA.Point, ReadDate, substationName) %>% 
  filter(row_number() == 1)

```


```{r Format_the_DataPowerFactor, include=FALSE}
# Ppivot the data for the pf calculation
# SCADA.Point contains, megawatts, megavars delievered, and megaVars received
pf_wide <- spread(pf_dedupe,SCADA.Point, ReadValue)

# rename the columns
pf_wide <- rename(pf_wide, 
                  mw=PMWD3D, 
                  mvar.delivered=PMQD3D, 
                  mvar.received=PMQD3R)

# Date Formating
# The data is interval ending. This means the dates need to be shifted 15 
# minutes back if we use the transformation below to avoid associating 
# the first 3 intervals with day the last interval for the day. 
#  
# Which is the next day at zero hundred hours (00:00))
# One way to solve this is convert readdate to as.Date( ) and then subtract 
# 15mins.
# Subtracting 15mins from the date will only work for intervals < than one hour
# Solved using lubridate, as.Date and POSXt was a pain.
pf_wide$dtReadDate <- parse_date_time(pf_wide$ReadDate, orders="mdy hm")
pf_wide$dtReadDate <- pf_wide$dtReadDate  - minutes(10)
pf_wide$dtReadDay  <- floor_date(pf_wide$dtReadDate,"day")
pf_wide$h <- hour(pf_wide$dtReadDate)

# force hour numeric, we will do the same for v_dedupe
# so we can join these columns on each other to join the 2 datasets
pf_wide$h <- as.numeric(pf_wide$h)

# Calculate the total MVARs MegaVARs
# mvar.received can be .na, when this occurs just use the mvar.delievered
# Set mvar to the difference (delv.-recev., if NO Received then use Delievered) 
pf_wide$mvar <- ifelse(!is.na(pf_wide$mvar.received), 
                       (pf_wide$mvar.delivered-pf_wide$mvar.received) 
                       ,pf_wide$mvar.delivered)

# Calculate the power factor using formuals below
# Power Factor = kw / (kw^2 + (Delieverd-Received)^2)  
# Power Factor = mw / (mw^2 + (mvar)^2)  
pf_wide$pf <- pf_wide$mw / 
    sqrt((pf_wide$mw * pf_wide$mw + pf_wide$mvar * pf_wide$mvar))

# Add the leading and lagging field to pf_wide.
# If mvar (+) Lagging, If mvar (-) Leading
pf_wide$desc <- ifelse(pf_wide$mvar < 0, 'Leading' , 'Lagging')

# Right shift the leading power factors for the chart
# Chart center is 1, and to show the directrion of the 
# phase angle we shift values to the right for the leading power factor.
pf_wide$pfChart <- ifelse( pf_wide$mvar < 0, (1-pf_wide$pf)+1 ,pf_wide$pf)

# Trim and Upper the Substation Name for Consistency
pf_wide$substationName <- trim(pf_wide$substationName)
pf_wide$substationName <- toupper(pf_wide$substationName)
```

```{r Format_the_Data_Voltage, include=FALSE}

# Clean and roll back 15min to accomodate interval ending and hours
v_dedupe$dtReadDate <- parse_date_time(v_dedupe$readdate, orders="mdy hm")
v_dedupe$dtReadDate  <- v_dedupe$dtReadDate  - minutes(15)
v_dedupe$dtReadDay  <- floor_date(v_dedupe$dtReadDate,"day")
v_dedupe$h <- hour(v_dedupe$dtReadDate)
# Create the hourmin 00:00 
v_dedupe$hm <- format(as.POSIXct(v_dedupe$readdate, format="%m/%d/%y %H:%M")
                      , format="%H:%M")

# force the value to numeric like we have for pf_wide$h
v_dedupe$h <- as.numeric(v_dedupe$h)

# divide all the votlages by 2 since we have 240v want to see 120v instead
# requirement: from customer
v_dedupe$Voltage <- v_dedupe$Voltage / 2

# Remove bad values over the HighLimit
v_dedupe <- subset(v_dedupe, v_dedupe$Voltage < voltageHighLimit)

# Trim and Upper the Substation Name for Consistency
v_dedupe$substationName <- trim(v_dedupe$substationName)
v_dedupe$substationName <- toupper(v_dedupe$substationName)

```


```{r Merge_the_Data, include=FALSE}

# create new dataframes with just the data we need to review
# to prepare to merge the two datasets so we can plot the bivarate charts
pf_col <- select(pf_wide,substationName,h,mvar ,pf, pfChart)
v_col <- select(v_dedupe, readdate,substationName,h, hm, Voltage)

# MERGE DATA FRAMES
# Join on substationName and hour.
# Voltage has more data and this will cause the  pf to be
# repeated for each additional 15min block.
# (How will this affect the summary of the data?)
# (Should I summarise to the hour and then join the two dataframes?)
# (And then do another summarise to the day?)
total <- left_join(v_col, pf_col, by= c("substationName", "h"))

```

```{r PowerFactorAnalysisAndShifting,include=FALSE}
# Second Attempt calcualte the min max from shifted values
totalSummary <- total %>%
  group_by(substationName)  %>%
  summarise(mean = mean(Voltage),
            vmedian = median(as.numeric(Voltage)),
            vmin = min(Voltage),
            vmax = max(Voltage),
            pfmean = mean(pfChart),
            pfmedian = median(as.numeric(pfChart)),
            pfmin = min(pfChart),
            pfmax = max(pfChart), 
            mvars = sum(mvar),
            n = n())  %>% 
  arrange(substationName)

# Add new fields to fix all the leading power factors where pf > 1
# Shift this value back to the original value
# for the table output summary
totalSummary_pretty <- totalSummary
totalSummary_pretty$pfmeanC <- ifelse(totalSummary_pretty$pfmean > 1, 
                                      (1-totalSummary_pretty$pfmean)+1,
                                      totalSummary_pretty$pfmean)

totalSummary_pretty$pfmedianC <- ifelse(totalSummary_pretty$pfmedian > 1, 
                                      (1-totalSummary_pretty$pfmedian)+1,
                                      totalSummary_pretty$pfmedian)

totalSummary_pretty$pfminC <- ifelse(totalSummary_pretty$pfmin > 1, 
                                      (1-totalSummary_pretty$pfmin)+1, 
                                     totalSummary_pretty$pfmin)

totalSummary_pretty$pfmaxC <- ifelse(totalSummary_pretty$pfmax > 1, 
                                      (1-totalSummary_pretty$pfmax)+1, 
                                     totalSummary_pretty$pfmax)

# Leading and Lagging Analysis
# Divide the lagging and leading dataframes into individual 
# data frames to help the user review the data in tables
leading <- subset(totalSummary_pretty, totalSummary_pretty$mvars <  0)
lagging <- subset(totalSummary_pretty, totalSummary_pretty$mvars >  0)

# Find the outliers in the data for leading and lagging
leading_outliers <- subset(leading, 
                    leading$pfmedian < pfLow |
                    leading$pfmedian > 1.02 |
                    leading$vmedian < voltageLow |
                    leading$vmedian > voltageHigh) 
                   
lagging_outliers <- subset(lagging, 
                   lagging$pfmedian < pfLow |
                   #lagging$pfmedian > pfHigh  |
                   lagging$vmedian  < voltageLow  |
                   lagging$vmedian  > voltageHigh )

```

```{r CleanUp_TableColumnNames, include=FALSE}
# Formatting for the summary output, pick the columns
# which have meaning for the users
total_review <- select(totalSummary_pretty,
                       substationName, 
                       mean,
                       vmedian,
                       vmin, 
                       vmax, 
                       pfmeanC,
                       pfmedianC,
                       pfminC,
                       pfmaxC,
                       mvars,
                       n)
# rename the columns
total_review <- rename(total_review, 
                       vmean=mean, 
                       pfmean=pfmeanC, 
                       pfmedian=pfmedianC, 
                       pfmin=pfminC, 
                       pfmax=pfmaxC, 
                       count=n)

```

## Voltage Lagging and Leading Daily

```{r SubVoltagePowerFactor,echo=FALSE,fig.width=12, fig.height=12, echo=FALSE, warning=FALSE }

# remove .na for easy plotting into ts - totalsummary
ts <-  subset(totalSummary_pretty, !is.na(totalSummary_pretty$pfmedian))

chart1 <- 
  ggplot(data=ts[!is.na(ts$pfmedian), ], 
       aes(x=pfmedian, y=vmedian)) +
    geom_point(size=3, color='green') + 
    geom_point(data=leading_outliers, 
               aes(x=pfmedian, y=vmedian), color='red', size=4) +
    geom_text(data=leading_outliers, 
              aes(label=substationName),hjust=0, vjust=2,size=3) +
    geom_errorbar(data=leading_outliers, 
                  aes(ymin = vmin, ymax = vmax), width = 0.0025, alpha=1/2) + 
    geom_point(data=lagging_outliers, 
               aes(x=pfmedian, y=vmedian), color='red', size=4) +
    geom_text(data=lagging_outliers, 
              aes(label=substationName),hjust=0, vjust=2,size=3) +
    geom_errorbar(data=lagging_outliers, 
                  aes(ymin = vmin, ymax = vmax), width = 0.0025, alpha=1/2) + 
    geom_errorbarh(data=leading_outliers, 
                   aes(xmax = pfmax, xmin = pfmin, height = .4), alpha=1/2) +
    geom_errorbarh(data=lagging_outliers, 
                   aes(xmax = pfmax, xmin = pfmin, height = .4), alpha=1/2) +
    scale_y_continuous(limits = c(110, 130)) +
    scale_x_continuous(limits = c(0.92, 1.10),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    geom_segment(aes(x = 0.98, y = voltageLow, xend = 0.98, 
                     yend = voltageHigh), size=0.15,colour='black') + # left
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                     yend = voltageHigh), size=0.15,colour='black') + # right
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                     yend = voltageHigh), size=0.05,colour='black') + # top
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                     yend = voltageLow), size=0.15,colour='black') + # bottom
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica')  +
    ggtitle("Substation Voltage & Power Factor" )

# remove all the outliers from the main dataset
ts <-  ts[!(ts$substationName %in% leading_outliers$substationName),]
ts <-  ts[!(ts$substationName %in% lagging_outliers$substationName),]

chart2 <- 
  ggplot(data=ts[!is.na(ts$pfmedian), ], 
       aes(x=pfmedian, y=vmedian,color=substationName)) +
    geom_point(size=3) + 
    geom_text(aes(label=substationName),hjust=-0.2, vjust=0.5,size=3) +
    geom_errorbar(aes(ymin = vmin, ymax = vmax), width = 0.0025,alpha=1/2) + 
    geom_errorbarh(aes(xmin = pfmin, xmax = pfmax,  height = .4),alpha=1/2) +
    scale_y_continuous(limits = c(110, 130)) +
    scale_x_continuous(limits = c(0.94, 1.06),
                       breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, 
                     yend = voltageHigh), size=0.15,colour='black') + # left
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                     yend = voltageHigh), size=0.15,colour='black') + # right
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                     yend = voltageHigh), size=0.05,colour='black') + # top
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                     yend = voltageLow), size=0.15,colour='black') + # bottom
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica') +
    #labs(linetype='custom title') +
    #scale_fill_continuous(guide = guide_legend()) +
    theme(legend.position="bottom")  +
    ggtitle("Substation Voltage & Power Factor" )

grid.arrange(chart1, arrangeGrob(chart2), ncol = 1)

```

Outliers Description:  
* Voltage Outlier (median) which will be above or below the high/low voltage  
* Power factor (median) which will left or right shift the substation on the chart.  

We do have to match the power factor to the hour and need to review if repeating the 
power factor for each interval in the hour avoids introduces anomolies. 

## Boxplot of Power Factor

```{r BoxPlotOfPowerFactor, echo=FALSE,fig.width=12, fig.height=10, echo=FALSE, warning=FALSE }
# reshape the y axis to fit the power factor use case / user expectations
ggplot(data=pf_wide[!is.na(pf_wide$pfChart), ], 
       aes(factor(substationName), pfChart)) +
    geom_boxplot(outlier.colour = "red") + 
    scale_y_continuous(limits = c(0.90, 1.10),
                        breaks=c(0.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                        labels=c(".92",".94", ".96", ".98","1", ".98", ".96",".94",".92")) +
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    ylab("power factor") + 
    geom_hline(yintercept = 1, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Power Factor by Substation" )

# calculate the boxplot's data for output
pfboxplotdata <- pf_wide %>% 
  group_by(substationName) %>% 
  summarise(min=min(pf,na.rm = TRUE), 
            Q1=quantile(pf, probs=0.25,na.rm = TRUE), 
            median=median(pf,na.rm = TRUE),
            mean=mean(pf,na.rm = TRUE),
            Q3=quantile(pf, probs=0.75,na.rm = TRUE),
            max=max(pf,na.rm = TRUE),
            count = n())  %>% 
        arrange(substationName)
```

### Boxplot Power Factor By Substation Data

```{r BoxPlotPowerFactorData, echo=FALSE}

kable(pfboxplotdata, digits=3)

```

This chart is used to show the shape of the distribution, its central value, 
and its variability. In a box and whisker plot: the ends of the box are the 
upper and lower quartiles, so the box spans the interquartile range. Rhe median 
is marked by a vertical line inside the box

The upper and lower "hinges" correspond to the first and third quartiles 
(the 25th and 75th percentiles). 

The upper whisker extends from the hinge to the highest value that is within 
1.5 * IQR of the hinge, where IQR is the inter-quartile range, or distance 
between the first and third quartiles. The lower whisker extends from the 
hinge to the lowest value within 1.5 * IQR of the hinge. Data beyond the end 
of the whiskers are outliers and plotted as points (as specified by Tukey).

* Outliers are in red.

## Histograms (Voltage & Power Factor)

```{r HistogramVoltagePowerFactor, echo=FALSE,fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}

# Histogram of voltage
p1<- ggplot(data = v_dedupe,aes(x = Voltage)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram (bin=0.5)")

# Histogram of power factor by lagging & leading
# need to avoid overlaying the bars on each other
p2<- ggplot(data=pf_wide[!is.na(pf_wide$pfChart), ]
            ,aes(x = pfChart, fill=desc)) + 
  geom_histogram(binwidth=0.002, color = 'black') + 
  theme_bw(base_family= 'Helvetica') + 
  theme(legend.position="bottom") + 
  theme(legend.position="none") +
  scale_x_continuous(limits = c(0.92, 1.08),
                       breaks=c(.92,.94,.96,.98,1,1.02,1.04,1.06,1.08),
                       labels=c("0.92",".94", ".96", ".98","1", ".98", ".96",".94","0.92")) +
  xlab("power factor") +
  ggtitle("Power Factor Histogram (bin=0.002)")

grid.arrange(p1, arrangeGrob(p2), ncol = 2)

```

A histogram is a bar graph of raw data that creates a picture of the data 
distribution. The bars represent the frequency of occurrence by classes of 
data. A histogram shows basic information about the data set, such as central 
location , width of spread , and shape.

### Lagging Substations 

```{r OutlierKTableNeg, echo=FALSE}
lagging_review <- total_review[
  (total_review$substationName %in% lagging_outliers$substationName),]

kable(lagging_review, digits=3)

```

### Leading Substations 

```{r OutlierKTablePos, echo=FALSE}
leading_review <- total_review[
  (total_review$substationName %in% leading_outliers$substationName),]

kable(leading_review, digits=3)

```

### All Substations

Need to pick a preference on what to display for the min and max and how to
distinguish when the min is lagging and max is leading, or when the substation
min and max spans the distance between the lagging and leading power factor.

```{r KTabletotalSummary, echo=FALSE}

kable(total_review, digits=2)

```

This table has multiple columns used to help right shift the charts and also 
the fields ending in 'C' which is the corrected value which is the 'original' 
value of the power factor without the leading shift.

Substations missing from the chart are missing data, either data for voltage or 
data needed to calculate power factor.  

TODO: An additional historical report of the daily min-max review of power factor
might be usefull when reporting leading and lagging at the daily level.  Or the 
number of times in the day when the substation switches from leading to lagging 
or lagging to leading.

## Voltage by 15min Intervals

```{r Voltage15MinuteIntervalsByHour, fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}

ggplot(data=v_dedupe, aes(x=hm, y=Voltage)) +
  geom_point(aes(color=substationName),
             position=position_jitter(h=1),
             alpha=1/2, 
             size=1) + 
  scale_y_continuous(limits = c(100, 130)) +
  scale_fill_brewer(palette="Set2") +
  geom_hline(yintercept=voltageLow) +
  geom_hline(yintercept=voltageHigh) +
  xlab("hr:min") +
  ggtitle("Voltage 15min Intervals Scatter Plot" ) +
  theme_bw(base_family= 'Helvetica') +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) +
  theme(axis.text.x = element_text(angle = -90, hjust = 0,size=6))  +
  theme(legend.position="bottom")

```

### Summary of Voltage for all data points.

```{r SummaryOfVoltageData, echo=FALSE, warning=FALSE}
#remove_max <- subset(total, total$Voltage < max(total$Voltage))
summary(v_dedupe$Voltage)
```

```{r VoltageBuckets, fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}
v_dedupe$voltage.bucket <- cut(v_dedupe$Voltage, 
                             breaks = c(0,110,115,120,125,130))

ggplot(data = v_dedupe, aes(x=h, y=Voltage))+
  geom_point(aes(color=voltage.bucket), size=1, position='jitter', alpha=.75)+
  scale_y_continuous()+
  scale_color_brewer(type="seq") +
  ggtitle("Voltage Intervals Grouped") +
  xlab("hour")+
  ylab("voltage") 
  #geom_line(aes(color=voltage.bucket), stat='summary', fun.y=median)
```

### Summary of Voltage Buckets
```{r SummaryOfVoltageBucketData, echo=FALSE, warning=FALSE}
#remove_max <- subset(total, total$Voltage < max(total$Voltage))
summary(v_dedupe$voltage.bucket)
```

Summary of Voltage for all data points excluding the max value for Lynchburg.

TODO: Another suggestion is that we can build groups of ranges and the number of 
incidents in those ranges is the factor to sort and search.

## Boxplot of Voltage

```{r BoxPlotOfVoltage, echo=FALSE,fig.width=12, fig.height=10, echo=FALSE, warning=FALSE }
# reshape the y axis to fit the power factor use case / user expectations
ggplot(data=v_dedupe[!is.na(v_dedupe$Voltage), ], 
       aes(factor(substationName), Voltage)) +
    geom_boxplot(outlier.colour = "red") + 
    theme_bw(base_family= 'Helvetica') +
    theme(legend.position="bottom") +
    xlab("") +
    geom_hline(yintercept = voltageHigh, color='blue') +
    geom_hline(yintercept = voltageLow, color='blue') +
    theme(axis.text.x=element_text(angle = -90, hjust = 0)) +
    ggtitle("Boxplot of Voltage by Substation" ) 

# http://www.statmethods.net/management/aggregate.html
voltageboxplotdata <- total %>% 
  group_by(substationName) %>% 
  summarise(min=min(Voltage,na.rm = TRUE), 
            Q1=quantile(Voltage, probs=0.25,na.rm = TRUE), 
            median=median(Voltage,na.rm = TRUE),
            mean=mean(Voltage,na.rm = TRUE),
            Q3=quantile(Voltage, probs=0.75,na.rm = TRUE),
            max=max(Voltage,na.rm = TRUE),
            count = n())  %>% 
        arrange(substationName)
```

### Boxplot Voltage By Substation Data

```{r BoxPlotVoltageData, echo=FALSE}

kable(voltageboxplotdata, digits=3)

```

This chart is used to show the shape of the distribution, its central value, 
and its variability. In a box and whisker plot: the ends of the box are the 
upper and lower quartiles, so the box spans the interquartile range. The median 
is marked by a vertical line inside the box

The upper and lower "hinges" correspond to the first and third quartiles 
(the 25th and 75th percentiles). 

The upper whisker extends from the hinge to the highest value that is within 
1.5 * IQR of the hinge, where IQR is the inter-quartile range, or distance 
between the first and third quartiles. The lower whisker extends from the hinge 
to the lowest value within 1.5 * IQR of the hinge. Data beyond the end of the 
whiskers are outliers and plotted as points (as specified by Tukey).

* Outliers are in red.

## Voltage & Power Factor Substation Outliers
Goal here is to see the movement of the substation voltage and power factor 
on the corrected chart displayed first in this analysis to see how the data 
moves in the regions for each individual interval of data.

### Substation Voltage & Power Factor Intervals

```{r SubstationVoltPFIntervals, echo=FALSE,fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}
# THIS ONE NEEDS TO BE total b/c it is both voltage and pf
ggplot(data=total[!is.na(total$pfChart), ], 
       aes(x=pfChart, y=Voltage)) +
    geom_point(aes(color=substationName),
               position=position_jitter(h=1),
               alpha=1/2, size=2) + 
    scale_fill_brewer(palette="Set2") +  
    scale_y_continuous(limits = c(100, 130)) +
    scale_x_continuous(limits = c(0.94, 1.06),
                       breaks=c(.94,.96,.98,1,1.02,1.04,1.06),
                       labels=c(".94", ".96", ".98","1", ".98", ".96",".94")) +
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 0.98, 
                     yend = voltageHigh), size=0.15,colour='black') + # left
    geom_segment(aes(x = 1.02, y = voltageLow,  xend = 1.02, 
                     yend = voltageHigh), size=0.15,colour='black') + # right
    geom_segment(aes(x = 0.98, y = voltageHigh, xend = 1.02, 
                     yend = voltageHigh), size=0.05,colour='black') + # top
    geom_segment(aes(x = 0.98, y = voltageLow,  xend = 1.02, 
                     yend = voltageLow), size=0.15,colour='black') + # bottom
    ggtitle("Substation Voltage & Power Factor" ) +
    ylab("median voltage") +
    xlab("median power factor") +  
    theme_bw(base_family= 'Helvetica') 

```

This chart represents the voltage and power factor for all data points 
collected for February 22, 2016. The chart allows one to see how much 
of the data falls out of the voltage and power factor defined region 
(0.98 (left) to 0.98 (right), and 114v to 128v).

We can see we have more points with Lagging power factors outside the 
required range for efficient power factor values.

## Histogram of Voltage Distribution (All Substations)

```{r HistogramOfVoltage, fig.width=12, fig.height=6, echo=FALSE, warning=FALSE}

h1 <- ggplot(data = v_dedupe,aes(x = Voltage)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram by Substation\n bin=0.5")

h2 <- ggplot(data = subset(v_dedupe,total$Voltage<117) ,aes(x = Voltage)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram by Substation (v < 117)\n bin=0.5")

h3 <- ggplot(data = subset(v_dedupe,total$Voltage>126) ,aes(x = Voltage),  binwidth=0.05) + 
  geom_histogram(binwidth=0.25, color = 'black', fill = '#333333') + 
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram by Substation (v > 126)\n bin=0.25")

grid.arrange(h1, arrangeGrob(h2, h3), ncol = 2)

```

A histogram is a bar graph of raw data that creates a picture of the data 
distribution. The bars represent the frequency of occurrence by classes of 
data. A histogram shows basic information about the data set, such as central 
location , width of spread , and shape.

## Substation Voltage Histograms
```{r VoltageHistchart, fig.width=10, fig.height=14, echo=FALSE, warning=FALSE}

ggplot(data = v_dedupe,aes(x = Voltage)) + 
  geom_histogram(binwidth=0.5, color = 'black', fill = '#333333') + 
  facet_wrap(~ substationName, ncol = 3) +
  theme_bw(base_family= 'Helvetica') +
  ggtitle("Voltage Histogram by Substation (bin=0.5)")

```

A histogram is a bar graph of raw data that creates a picture of the data 
distribution. The bars represent the frequency of occurrence by classes of 
data. A histogram shows basic information about the data set, such as central 
location , width of spread , and shape.

Would it be benefical to have hitogram by day for a single substation over 
a time period?  Maybe histogram per month?


## Substation Voltage by Time (intervals)

```{r VoltageScatterchart, fig.width=10, fig.height=14, echo=FALSE, warning=FALSE}

ggplot(data=v_dedupe, aes(x=hm, y=Voltage)) +
  geom_point(alpha=1/5) + 
  facet_wrap(~ substationName, ncol = 3) +
  xlab("hr:min") +
  theme_bw(base_family= 'Helvetica')  +
  theme(axis.text.x = element_blank())

```

The scatter plot shows the data along the time axis for the intervals for the 
day. The interesting point in this chart, which is similar to the histogram is 
how the shading changes from dark to grey, which is the points stacking on top 
of each other.  

It takes 5 points on top of each other to make a solid point on this chart.  
This demonstrates how the data is spread out over the ranges through the day 
by hour.

## Additional Charts for exploratory review.

### MegaWatt By Substation

```{r MeagWattsScatterPlot, fig.width=10, fig.height=8, echo=FALSE, warning=FALSE}

ggplot(data=pf_wide[!is.na(pf_wide$mw), ], aes(x=h, y=mw)) +
  geom_point() + 
  facet_wrap(~substationName, scales='free', ncol=3) +
  ggtitle("Mega Watt Scatter Plot by Substation" ) +
  theme_bw(base_family= 'Helvetica') 

```

### Power Factor by Substation and Hour 

```{r PowerFactorChart voltage, fig.width=10, fig.height=8,echo=FALSE, warning=FALSE}

ggplot(aes(x = h, y = pf, color=desc), data = pf_wide) + 
  geom_point(size = 2, position = 'jitter') +
  facet_wrap(~substationName) + 
  ggtitle('Power Factor (Leading and Lagging)') +
  theme_bw(base_family= 'Helvetica') 

```

Leading is colored instead of right shifted in these charts.

### Power Factor Summary By Substation

```{r PowerFactorSummarySort voltage, echo=FALSE}
# sort the data
sorted_select <- pf_wide[order(pf_wide$substationName, 
                                     pf_wide$dtReadDate),] 

sorted_review <- select(sorted_select,
                      dtReadDate,
                      substationName,
                      mw,
                      mvar.delivered, 
                      mvar.received, 
                      pf, 
                      desc)

sorted_review <- sorted_select[,c("dtReadDate",
                                  "substationName",
                                  "mw",
                                  "mvar.delivered",
                                  "mvar.received",
                                  "pf",
                                  "desc")]

kable(sorted_review, digits=2)

```

### MegaVARS By Substation

```{r MegaVarsScatterPlot, fig.width=10, fig.height=8, echo=FALSE, warning=FALSE}

# ggplot(pf_wide, aes(x=h, y = value, color = variable)) + 
#     geom_point(aes(y = mvar.received, col = "Received")) + 
#     geom_point(aes(y = mvar.delivered, col = "Delivered")) +
#     facet_wrap(~substationName, scales='free', ncol=3)


# ggplot(data=pf_wide[!is.na(pf_wide$mvar.received), ],
#        aes(x=h, y=mvar.received)) +
#   facet_wrap(~substationName, scales='free', ncol=3) +
#   ggtitle("Mega VARs Received by Substation" ) +
#   xlab("hour") +
#   theme_bw(base_family= 'Helvetica')
# 
# ggplot(data=pf_wide[!is.na(pf_wide$mvar.delivered), ],
#        aes(x=h, y=mvar.delivered)) +
#   facet_wrap(~substationName, scales='free', ncol=3) +
#   ggtitle("Mega VARs Delivered by Substation" ) +
#   xlab("hour") +
#   theme_bw(base_family= 'Helvetica')

```

## References

### Bad Values
Any voltage over 180v is considered bad data, or a spike and the data will be 
removed.

### Interval Ending
Votlage Data

### Chapel Hill Power Factor 
Chapel Hill has a different power meter.  It does not register MVARS received 
in a separate register.  It is (+) if delivered and (–) if received.  Therefore, 
pf will have to be calculated with just the one number for MVARS. 

### Votlage Calculation
Voltage is being divided by 2 to adjust the view to 120 volts since the data 
being collected is 240 volts.

### Power Factor  
Instructions regarding PF calculation:  
https://en.wikipedia.org/wiki/Power_factor  

http://www.slideshare.net/RossWelshVA/power-factor-correction-the-easiest-biggest-green-initiative

Q, Reactive Power - kVAR -> PMQD3D Delivered, PMQR3D Received  
  (PMQD3D-PMQR3D)  total VARs  
S, Apparent Power    
P, Real Power - kW -> PMWD3D (MW)

S^2 = P^2 + Q^2  
S   = SQRT(p^2 + Q^2)  

Power Factor = (P, Real Power) / (S, Apparent Power)  
Power Factor = kw / (kw^2 + kvar^2)  
Power Factor = kw / (kw^2 + (Delieverd-Received)^2)  

### Would it be nice to see the trend of kw or kvar increasing or decreasing 
to predict when power factor is on a downward trend.... 

### Leading and Lagging Power Factor
* Delivered vars - Received vars   > 0 Lagging  
* Delivered vars - Received vars   < 0 Leading

Summary of notes to be reviewed:

The original definition of MAX PF:  

Least positive Lagging PF value for the given set of data.    

A Lagging PF is defined as any value where PMQD3D – PMQD3R is greater than zero.  
Because PMQD3D is all zeroes for BOYNTON and CULLEOKA for the samples below, 
there is never a Lagging PF value for the date.  Therefore, the MAX PF would 
be zero for the given day.  

From the discussion:   
· Min PF - Least negative (Leading) PF value for the given subset  
· Max PF - Least positive (Lagging) PF value for the given subset  
· Avg PF - Most common value for the given subset OR Mean Value  

Notes:
Phase 1 - is just charting and quick review of the data.  
Phase 2 - will apply statistical models and analysis if required.

TODO: This is a daily analysis so the data is available after the day is over.  
To improve this type of analysis we need to see patterns and trends which occur 
which can help predict an event which is about to happen. 

http://vita.had.co.nz/papers/tidy-data.pdf

https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf
https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf
http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/
http://www.statmethods.net/management/aggregate.html